---
title: "Clustering PA"
author: "Conor Tompkins"
date: "August 12, 2018"
output: html_document
runtime: shiny
---
#Clustering with PCA and Kmeans
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE, 
                      warning = FALSE)
```

In this post I will use the census API discussed in [the last post]() to cluster the Allegheny County census tracts using PCA and k-means.

#Setup
```{r}
library(tidyverse)
library(tidycensus)
library(tigris)
library(sf)
library(broom)
library(ggfortify)
library(viridis)
library(janitor)
library(scales)
library(ggthemes)

options(tigris_use_cache = TRUE)
census_api_key("a16f3636406d2b544871e2ae49bb318c3ddcacba")

theme_set(theme_minimal())

census_vars <- load_variables(2010, "sf1", cache = TRUE)
```

#Download data
This code downloads data about the ethnicities resident in the tracts.
```{r}
vars_demo <- c(white = "P0050003", 
          black = "P0050004", 
          asian = "P0050006", 
          hispanic = "P0040003")
#age vars men and women
#P0120003:P0120049

get_decennial(geography = "tract", 
                           variables = vars_demo,
                           state = "PA", 
                           geometry = FALSE,
                           summary_var = "P0010001") %>% 
  arrange(GEOID) %>% 
  mutate(value = value / summary_value) %>% 
  select(-summary_value) %>% 
  spread(variable, value) %>% 
  rename_at(vars("white", "black", "asian", "hispanic"), funs(str_c("percent_", .))) -> allegheny_demographics

allegheny_demographics <- replace(allegheny_demographics, is.na(allegheny_demographics), 0)
```

This code downloads information about the housing stock in each tract, specifically what %  of housing units are owned outright, owned with a loan, or rented.
```{r}
vars_housing <- c(units_owned_loan = "H0110002",
          units_owned_entire = "H0110003",
          units_rented = "H0110004")

get_decennial(geography = "tract", 
                           variables = vars_housing,
                           state = "PA", 
                           geometry = FALSE,
                           summary_var = "H0110001") %>% 
  arrange(GEOID) %>% 
  mutate(value = value / summary_value) %>% 
  select(-summary_value) %>% 
  spread(variable, value) %>% 
  rename_at(vars("units_owned_entire", "units_owned_loan", "units_rented"), funs(str_c("percent_", .))) -> allegheny_housing

allegheny_housing <- replace(allegheny_housing, is.na(allegheny_housing), 0)
```

This code requests the total population of each tract.
```{r}
#originally I used age-sex variables, but they were not useful
vars_age_total <- census_vars %>% 
  filter(concept == "P12. Sex By Age [49]",
         label != "Total population",
         label != "Male:")

get_decennial(geography = "tract", 
                           variables = vars_age_total$name,
                           state = "PA", 
                           geometry = FALSE,
                           summary_var = "P0120001") %>% 
  rename(var_id = variable) %>% 
  mutate(value = value / summary_value) %>% 
  spread(var_id, value) -> allegheny_age_sex

colnames(allegheny_age_sex) <- c("GEOID", "NAME", "summary_value", vars_age_total$label)

allegheny_age_sex %>% 
  clean_names() %>% 
  rename(GEOID = geoid,
         NAME = name,
         total_population = summary_value) -> allegheny_age_sex

allegheny_age_sex <- replace(allegheny_age_sex, is.na(allegheny_age_sex), 0)

allegheny_age_sex %>% 
  select(GEOID, NAME, total_population) -> allegheny_age_sex
```

This code requests the geometry of each tract that I will use to map them later.
```{r}
get_decennial(geography = "tract", 
                           variables = vars_housing,
                           state = "PA", 
                           geometry = TRUE) %>% 
  select(-c(variable, value)) %>% 
  mutate(id = str_c(GEOID, NAME, sep = " | ")) -> allegheny_geo
```

This joins the 4 dataframes together.
```{r}
allegheny_geo %>% 
  left_join(allegheny_housing) %>% 
  left_join(allegheny_demographics) %>% 
  left_join(allegheny_age_sex) %>% 
  #mutate(id = str_c(GEOID, NAME, sep = " | ")) %>% 
  distinct() -> allegheny
```
#Exploratory graph
This graph compares the percent of white residents to the remaining variables in the data. It is already obvious that percent_white and percent_black are negatively correlated with each other.
```{r}
allegheny %>%
  st_set_geometry(NULL) %>% 
  select(contains("percent"), total_population) %>% 
  gather(variable, value, -percent_white) %>% 
  ggplot(aes(percent_white, value)) +
  geom_point(alpha = .5) +
  geom_smooth() +
  facet_wrap(~variable, scales = "free")
```

#Prepare for PCA
This code prepares the data for PCA:
```{r}
allegheny %>%
  select(-c(NAME, GEOID)) %>% 
  st_set_geometry(NULL) %>% 
  remove_rownames() %>% 
  column_to_rownames(var = "id") -> allegheny_pca


allegheny_pca %>% 
  prcomp(scale = TRUE) -> pc
```

```{r include=FALSE}
pc %>% 
  tidy() %>% 
  head()

pc %>% 
  tidy("pcs")

pc %>% 
  augment(data = allegheny_pca) %>% 
  as_tibble() -> df_au

df_au %>% 
  head()
```
This shows how the PCs explain the variance in the data. As explained earlier, the first few PCs explain most of the variance in the data.
```{r}
pc %>% 
  tidy("pcs") %>%
  select(-std.dev) %>% 
  gather(measure, value, -PC) %>% 
  mutate(measure = case_when(measure == "percent" ~ "Percent",
                             measure == "cumulative" ~ "Cumulative")) %>% 
    ggplot(aes(PC, value)) +
    geom_line() +
    geom_point() +
    facet_wrap(~measure) +
    labs(title = "Variance explained by each principal component",
         x = "Principal Component",
         y = NULL) +
    scale_x_continuous(breaks = 1:8) +
    scale_y_continuous(label = percent)
```

This shows how the PCA function rearranged the data to maximize the variance in the first few PCs. PC1 is largely defined by the percent of a tract that is white or black, the percent of housing units that are owned, and the total population of the tract. The "percent_white" and "percent_black" arrows point in opposite directions, which reflects Pittsburgh's status as a segregated city. LINKS

PC2 explains less of the variance, and is influenced by the total population of a tract, and the percent of a tract that is hispanic, asian, or black.
```{r}
allegheny %>% 
  select(-c(NAME, GEOID)) %>% 
  st_set_geometry(NULL) %>% 
  nest() %>% 
  mutate(pca = map(data, ~ prcomp(.x %>% select(-id), 
                                  center = TRUE, scale = TRUE)),
         pca_aug = map2(pca, data, ~augment(.x, data = .y))) -> allegheny_pca2

allegheny_pca2 %>% 
mutate(
    pca_graph = map2(
      .x = pca,
      .y = data,
      ~ autoplot(.x, loadings = TRUE, loadings.label = TRUE,
                 loadings.label.repel = TRUE,
                 data = .y) +
        theme_bw() +
        labs(x = "Principal Component 1",
             y = "Principal Component 2",
             title = "First two principal components of PCA on Allegheny County Census data")
    )
  ) %>%
  pull(pca_graph)
```

This code maps the first two PCs to the tracts. 
```{r}
df_au %>% 
  rename(id = .rownames) %>% 
  gather(variable, value, -c(id)) -> df_au_long

#st_erase <- function(x, y) {
#  st_difference(x, st_union(st_combine(y)))
#}

#allegheny_water <- area_water("PA", "Allegheny", class = "sf")

#allegheny_geo <- st_erase(allegheny_geo, allegheny_water)


allegheny_geo %>% 
  left_join(df_au %>% rename(id = .rownames)) %>% 
  select(id, contains(".fitted")) %>% 
  #gather(variable, value, contains("percent")) %>% 
  gather(pc, pc_value, contains(".fitted")) %>% 
  mutate(pc = str_replace(pc, ".fitted", "")) -> allegheny_pca_map

allegheny_pca_map %>% 
  filter(pc %in% c("PC1", "PC2")) %>% 
  ggplot(aes(fill = pc_value, color = pc_value)) + 
  geom_sf() + 
  facet_wrap(~pc) +
  #coord_sf(crs = 26911) + 
  scale_fill_viridis("Principal component value", option = "magma") + 
  scale_color_viridis("Principal component value", option = "magma") +
  labs(title = "Allegheny County",
       subtitle = "American Community Survey") +
  theme(axis.text = element_blank())
```


#k-means
Next I will use k-means to cluster the PC data.
```{r}
df_au_long %>% 
  filter(str_detect(variable, "PC")) %>% 
  spread(variable, value) %>% 
  select(-id) -> allegheny_kmeans
```

This code clusters the data using 1 to 9 clusters.
```{r}
kclusts <- tibble(k = 1:9) %>%
  mutate(
    kclust = map(k, ~kmeans(allegheny_kmeans, .x)),
    tidied = map(kclust, tidy),
    glanced = map(kclust, glance),
    augmented = map(kclust, augment, allegheny_kmeans)
  )
```

```{r}
clusters <- kclusts %>%
  unnest(tidied)

assignments <- kclusts %>% 
  unnest(augmented)

clusterings <- kclusts %>%
  unnest(glanced, .drop = TRUE)
```

Based on this "elbow chart", the optimum number of clusters is most likely 2.
```{r}
ggplot(clusterings, aes(k, tot.withinss)) +
  geom_line() +
  geom_point() +
  geom_vline(xintercept = 2, linetype = 2) +
  scale_x_continuous(breaks = 1:9) +
  labs(x = "Number of clusters",
       y = "Between-cluster sum of squares")
```

We can visualize how the data would look if it were assigned to a different number of clusters. Clearly the clustering algorithm experiences diminishing returns after 2 or 3 clusters.
```{r}
ggplot(assignments, aes(.fittedPC1, .fittedPC2)) +
  geom_point(aes(color = .cluster), alpha = .7) + 
  facet_wrap(~ k) +
  scale_color_discrete("Cluster") +
  labs(x = "PC1",
       y = "PC2")
```

This code divides the data into 2 clusters and maps the clusters onto the tract map.
```{r}
df_au_long %>% 
  filter(str_detect(variable, "PC")) %>% 
  spread(variable, value) %>% 
  select(-id) -> allegheny_kmeans

kclust <- kmeans(allegheny_kmeans, centers = 2)

kclust %>% 
  augment(df_au_long %>% 
    filter(str_detect(variable, ".fitted")) %>% 
    spread(variable, value)) -> allegheny_kmeans

get_decennial(geography = "tract", 
                           variables = vars_housing,
                           state = "PA", 
                           geometry = TRUE) %>% 
  select(-c(variable, value)) %>% 
  mutate(id = str_c(GEOID, NAME, sep = " | ")) -> allegheny_geo

allegheny_geo %>% 
  left_join(allegheny_kmeans) -> allegheny
```

```{r}
#st_erase <- function(x, y) {
#  st_difference(x, st_union(st_combine(y)))
#}

#allegheny_water <- area_water("PA", "Allegheny", class = "sf")

#allegheny <- st_erase(allegheny, allegheny_water)

allegheny %>% 
  ggplot(aes(fill = .cluster, color = .cluster)) +
  geom_sf() +
  scale_fill_viridis("Cluster", discrete = TRUE) + 
  scale_color_viridis("Cluster", discrete = TRUE) +
  theme(axis.text = element_blank())
```